ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-02-26 pyked.tex

%====================================================================
\section{Overview of format}\label{sec:overview-of-format}

The \ck{} file is a YAML-format file. This format offers the advantages of being
human readable, written in plain text, and having parsers in most common
programming languages, including Python, C++, Java, Perl, MATLAB, and many
others. The YAML syntax is quite simple. The basic file structure is made of
mappings, delimited by a colon. The key for the mapping is on the left of the
colon, and the value on the right can be a single value, a sequence of values,
a nested mapping, or some combination of these:

\begin{minted}{yaml}
    key1: value  # Single-value mapping
    key2: [0, 1, 2]  # Mapping with sequence
    key3:  # Alternative sequence format
      - value
      - value
    key4:  # Nested mapping
      key5: 0
\end{minted}

The \yaml|value| can be a string, integer, or floating point number. The type is
inferred from the \yaml|key| when the value is read. The \ck{} format is
designed to include all of the information necessary to simulate a given
experiment and can be broken into two main sections. The first section details
all of the ``meta'' information about the experiment and the \ck{} file. In this
section, the common \yaml|author| mapping is used for any fields where an author
must be specified. The \yaml|author| mapping contains the following fields:
%
\begin{itemize}
    \item \yaml|name| (required, string): The author's full name
    \item \yaml|ORCID| (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
\todo[author=Kyle,inline]{one difference ReSpecTh called out from PrIMe is that
the latter added a copyright line in all files, claiming copyright on the data...
do we want to be incendiary and point out that you can't copyright data?}
%
\begin{itemize}
    \item \yaml|file-author| (required, \yaml|author|-type mapping): The author
    of the \ck{} file
    \item \yaml|file-version| (required, integer): The version of the \ck{} file
    \item \yaml|chemked-version| (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yaml|reference| (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yaml|authors| (required, sequence): A sequence of \yaml|author|
        mappings
        \item \yaml|journal| (required, string): The name of the publishing
        journal
        \item \yaml|year| (required, integer): The year of publication
    \end{itemize}
    \item \yaml|experiment-type| (required, string): One of
    \yaml|ignition delay| or \yaml|flame speed|
    \item \yaml|apparatus| (required, string): The type of apparatus used to
    perform the experiment
\end{itemize}

In addition to the meta information, the data encoded in the \ck{} file is in
a mapping called \yaml|datapoints|. The \yaml|datapoints| mapping contains a
sequence of nested mappings, with the following information for each element of
the sequence:
%
\begin{itemize}
    \item \yaml|temperature| (required, sequence): The temperature of the
    experiment, with units
    \item \yaml|ignition-delay| (optional, sequence): The ignition delay of the
    experiment, with units, if the \yaml|experiment-type| is
    \yaml|ignition delay|
    \item \yaml|flame-speed| (optional, sequence): The flame speed of the
    experiment, with units, if the \yaml|experiment-type| is \yaml|flame speed|
    \item \yaml|pressure| (required, string): The pressure of the experiment,
    with units
    \item \yaml|composition| (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yaml|species-name| (required, string): The name of the species
        \item \yaml|mole-fraction|, \yaml|mass-fraction|, \yaml|mole-percent|
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yaml|composition|
        sequence.
        \item \yaml|InChI|, \yaml|SMILES|, \yaml|atomic-composition|, or
        \yaml|elemental-composition| (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yaml|atomic-composition| and \yaml|elemental-composition|
        are synonyms) as a sequence
    \end{itemize}
    \item \yaml|ignition-type| (required, mapping); The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yaml|type| (required, string): How the ignition delay was
        measured, one of \yaml|d/dt max| to indicate the ignition point is
        found at the maximum of the time derivative of the \yaml|target|, or
        \yaml|max| or \yaml|min| to indicate the ignition point is at the
        maximum or minimum of the \yaml|target|
        \item \yaml|target| (required, string): The target for the type of the
        ignition point, one of \yaml|temperature|, \yaml|pressure|, \yaml|OH|,
        \yaml|OH*|, \yaml|CH|, or \yaml|CH*|
    \end{itemize}
\end{itemize}

Each of the quantities in a datapoint must be specified with units. This should
be done as a single string associated with the first element of the sequence.
The units of the quantity are validated to ensure they have the appropriate
dimensionality for the quanity. In addition, each of the quantities in the
\yaml|datapoint| can optionally be assigned an uncertainty. This uncertainty can
be either absolute or relative, and is specified as elements of the sequence of
the associated key. For example, the absolute uncertainty of the temperature and
the relative uncertainty of the ignition delay might be specified as follows:

\begin{minted}{yaml}
datapoints:
  - temperature:
      - 1100 kelvin
      - uncertainty-type: absolute
        uncertainty: 10 kelvin
    ignition-delay:
      - 10 us
      - uncertainty-type: relative
        uncertainty: 0.1
    ...
\end{minted}

%====================================================================
\section{PyKED architecture}
%
\pk{} is a Python package that currently provides the reference implementation
of the interface to \ck{} files. \pk{} reads \ck{} files, validates their
structure and content, and allows the user to interact with the datapoints
contained in the \ck{} file.

The basic user interface to the \ck{} file in \pk{} is through the
\python|ChemKED| class. The \python|ChemKED| class constructor takes the file
name of a \ck{} file or a Python dictionary containing the contents of a \ck{}
file as its argument. When the file or dictionary is loaded, it is validated to
have the correct format by Cerberus \autocite{cerberus}.

The fields of the \ck{} file are stored as instance attributes of the
\python|ChemKED| class. The following attributes are available with the
appropriate data:
%
\begin{itemize}
    \item \python|chemked_version|, \python|file_version|, \python|file_author|,
    \python|experiment_type|: Stores the values from the \ck{} file
    \item \python|reference|: Instance of a \python|namedtuple| containing all
    the information from the literature reference associated with the data
    \item \python|apparatus|: Instance of a \python|namedtuple| containing all
    the information about the apparatus used to perform the experiment
    \item \python|datapoints|: A Python list of \python|DataPoint| instances
\end{itemize}

The \python|DataPoint| class stores the information associated with a single
data point in the \ck{} file. Similar to the \python|ChemKED| class, the
information from the \ck{} file are stored as attributes of class instances,
including
%
\begin{itemize}
    \item \python|equivalence_ratio|: The value of the equivalence ratio, if
    present. For informational purposes only, no validation is done of the value
    \item \python|ignition_delay|, \python|temperature|, \python|pressure|,
    \python|pressure_rise|: The value of these data are stored as instances of
    the \python|Quantity| class from the Pint \autocite{Grecco2016} package,
    possibly with associated uncertainty
    \item \python|composition|: A list of dictionaries of the species and their
    respective amounts. The values are validated so that \yaml|mole-percent|,
    \yaml|mole-fraction|, or \yaml|mass-fraction| cannot be mixed for a single
    data point, and so that the sum of the values is approximately 1.0, or
    100.0 for \yaml|mole-percent|.
    \item \python|composition_type|: A string indicating the type of composition
    information for the data point---one of \python|'mole-percent'|,
    \python|'mole-fraction'|, or \python|'mass-fraction'|
    \item \python|volume_history|: If the \yaml|volume-history| of an RCM
    experiment is provided in the \ck{} file, it is stored in this attribute as
    a \python|namedtuple|, and the actual values are stored in NumPy arrays
    \autocite{vanderWalt:2011np}.
\end{itemize}

The \python|DataPoint| class has two instance methods defined, called
\python|get_cantera_mole_fraction| and \python|get_cantera_mass_fraction|. With
these methods, the composition of the reactant mixture can be output to a format
that can be used to set the composition of a Cantera \python|Solution|. Note
that the \python|composition| specification is not required to contain the
molecular weights of the components, so calling
\python|get_cantera_mass_fraction| when \yaml|mole-percent| or
\yaml|mole-fraction| has been used in the \ck{} file will raise a
\python|ValueError| (and vice versa for \python|get_cantera_mole_fraction|).

The \python|ChemKED| class has one instance method defined, called
\python|get_dataframe|. Using this method, the user can get an instance of a
Pandas~\autocite{pandas} \python|DataFrame| that contains the information in the
list of \python|DataPoint|s. The user can specify the columns provided in the
\python|DataFrame| by passing a list of column names to the
\python|output_columns| argument of the \python|get_dataframe| function. The
possible columns are not case-sensitive and are:

\noindent\begin{itemize*}
    \item \python|'Temperature'|
    \item \python|'Pressure'|
    \item \python|'Ignition Delay'|
    \item \python|'Composition'|
    \item \python|'Equivalence Ratio'|
    \item \python|'Reference'|
    \item \python|'Apparatus'|
    \item \python|'Experiment Type'|
    \item \python|'File Author'|
    \item \python|'File Version'|
    \item \python|'ChemKED Version'|
\end{itemize*}

In addition, specific fields from the Reference and Apparatus attributes can
be included by specifying the name after a colon. These options are:

\noindent\begin{itemize*}
    \item \python|'Reference:Volume'|
    \item \python|'Reference:Journal'|
    \item \python|'Reference:DOI'|
    \item \python|'Reference:Authors'|
    \item \python|'Reference:Detail'|
    \item \python|'Reference:Year'|
    \item \python|'Reference:Pages'|
    \item \python|'Apparatus:Kind'|
\end{itemize*}

% This weirdness is necessary to prevent the list from running over the margin
\noindent\begin{itemize*}
    \item \python|'Apparatus:Facility'|
    \item \python|'Apparatus:Institution'|
\end{itemize*}

Only the first author is printed when \python|Reference| or
\python|Reference:Authors| is selected because the whole author list may be
quite long.

@@

The \ck{} file that will be used in this example can be found in the
\python|tests| directory of the PyKED repository,
\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself
%
\begin{minted}{yaml}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
fake-key: 0 # Again here because Atom's syntax highlighting is busted :(
\end{minted}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the data point is from the work of \textcite{Mittal2006a}:
%
\begin{minted}{yaml}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
  fake-key: 0 # Again here because Atom's syntax highlighting is busted :(
\end{minted}

Finally, there is just a single datapoint in this file, which codes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure,
%
\begin{minted}{yaml}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        - 0 # Dammit Atom
        ...
\end{minted}

The values for the \yaml|volume-history| are truncated to save space. The user
can load this file into \pk{} and run a simulation in Cantera with the following
code:

\begin{minted}{python}
import cantera as ct
from pyked import ChemKED

ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]

T_0 = dp.temperature.to('K').magnitude
P_0 = dp.pressure.to('Pa').magnitude
X_0 = dp.get_cantera_mole_fraction()

gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_0, P_0, X_0

reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))

time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

netw = ct.ReactorNet([reac])
while netw.time < 0.05:
    netw.step()

\end{minted}

@@

The \ck{} file used in this example can be found online.\todo{add url}{}
The data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol.
Examining the file, we first find the section that specifies the information about the \ck{} file itself:
%
\begin{minted}{yaml}
file-author:
  name: Morgan Mayer
  ORCID: 0000-0001-7137-5721
file-version: (0,1)
chemked-version: 0.0.1
screw-you-atom: 0
\end{minted}
%
Then the reference information:
%
\begin{minted}{yaml}
reference:
  doi: 10.1016/j.combustflame.2011.08.014
  authors:
    - name: Ivo Stranic
    - name: Deanna P. Chase
    - name: Joseph T. Harmon
    - name: Sheng Yang
    - name: David F. Davidson
    - name: Ronald K. Hanson
  journal: Combustion and Flame
  year: 2011
  volume: 159
  pages: 516-527
  detail:
experiment-type: ignition delay
apparatus:
  kind: shock tube
  institution: High Temperature Gasdynamics Laboratory, Stanford University
  facility: stainless steel shock tube
dammit-atom: 0
\end{minted}
%
In this \ck{} file, multiple data points are specified with some common
conditions, including a common mixture composition and common defintion of
ignition delay. Therefore, a \yaml|common-properties| section is specified.
%
\begin{minted}{yaml}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
atom-is-bad-at-syntax-highlighting: 0
\end{minted}
%
This block utilizes the ability of YAML files to define an anchor using the
\yaml|&| symbol and refer to that section later using the \yaml|*| symbol, as
can be seen in the definition of the \yaml|datapoints| section:
%
\begin{minted}{yaml}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{minted}

In this example, we would like to run constant volume simulations at each of the
pressure and temperature conditions in the \yaml|datapoints| list.

\begin{minted}{python}
import cantera as ct
from pyked import ChemKED

@@

X_0 = ck.datapoints[0].get_cantera_mole_fraction()

for dp in ck.datapoints:
    T_0 = dp.temperature.to('K').magnitude
    P_0 = dp.pressure.to('Pa').magnitude
    gas.TPX = T_0, P_0, X_0
    reac = ct.IdealGasReactor(gas)
    netw = ct.ReactorNet([reac])
    while reac.T < T_0 + 400:
        netw.step()

    print('The ignition delay for T_0={}, P_0={} is: {} seconds'.format(
        T_0, P_0, netw.time
    ))
\end{minted}
%====================================================================
\section{Conclusions and future work}
%
\todo[inline]{both to do}

In this article, we introduced the \ck{} data format for describing measurements
taken from fundamental combustion experiments, as well as the Python-based tool
\pk{} for working with the format. Currently, \ck{} supports ignition delay data
from studies using shock tubes and rapid compression machines.

We are currently incorporating support for specifying uncertainty in all quantities.
Version 1.0 of \ck{}\slash \pk{} will include ...
Later versions will support measurements from additional fundamental experiments,
including laminar flames and flow reactors.

>>>>

%====================================================================
\section{Overview of ChemKED format}\label{sec:overview-of-format}

This section will provide a brief introduction to the \ck{} database format.
This is not a complete listing of available fields and more complete examples of
\ck{} files can be found in the \pybox{tests} directory of the \pk{} package. In
addition, in Section~\ref{sec:usage-example}, we will provide examples of \ck{}
files. Note that in the following, text with gray background is a \yabox{yaml}
keyword or value, while text with a yellow background is \pybox{Python} code.

\ck{} files use the YAML data serialization format. This format offers the
advantages of being human readable, written in plain text, and having parsers in
most common programming languages, including Python, C++, Java, Perl, MATLAB,
and many others. The YAML syntax is quite simple. The basic file structure is
made of mappings, delimited by a colon. The key for the mapping is on the left
of the colon, and the value on the right can be a single value, a sequence of
values, a nested mapping, or some combination of these:
%
\begin{yamlbox}
key1: value  # Single-value mapping
key2:  # Sequence format
  - value
  - value
key3:  # Nested mapping
  key4: 0
key5:  # Sequence of mappings
  - key-s1: 0
    key-s2: value
  - key-s2: value
    key-s1: 0
\end{yamlbox}

The \yabox{value} can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox{author} mapping is used for any fields where
an author must be specified. The \yabox{author} mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox{name} (required, string): The author's full name
    \item \yabox{ORCID} (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
\todo[author=Kyle,inline]{one difference ReSpecTh called out from PrIMe is that
the latter added a copyright line in all files, claiming copyright on the data...
do we want to be incendiary and point out that you can't copyright data?}
%
\begin{itemize}
    \item \yabox{file-author} (required, \yabox{author}-type mapping): The
    author of the \ck{} file
    \item \yabox{file-version} (required, integer): The version of the \ck{}
    file
    \item \yabox{chemked-version} (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox{reference} (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox{authors} (required, sequence): A sequence of \yabox{author}
        mappings
        \item \yabox{journal} (required, string): The name of the publishing
        journal
        \item \yabox{year} (required, integer): The year of publication
    \end{itemize}
    \item \yabox{experiment-type} (required, string): Currently, only
    \yabox{ignition delay} is supported
    \item \yabox{apparatus} (required, string): The type of apparatus used to
    perform the experiment
\end{itemize}

In the \ck{} file, the experimental data is encoded in a sequence of mappings,
with the following information required in each element of the sequence:
%
\begin{itemize}
    \item \yabox{temperature} (required, sequence): The temperature of the
    experiment, with units
    \item \yabox{ignition-delay} (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox{pressure} (required, string): The pressure of the experiment,
    with units
    \item \yabox{composition} (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox{species-name} (required, string): The name of the species
        \item \yabox{mole-fraction}, \yabox{mass-fraction}, \yabox{mole-percent}
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox{composition}
        sequence.
        \item \yabox{InChI}, \yabox{SMILES}, \yabox{atomic-composition}, or
        \yabox{elemental-composition} (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox{atomic-composition} and
        \yabox{elemental-composition} are synonyms) as a sequence
    \end{itemize}
    \item \yabox{ignition-type} (required, mapping); The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox{type} (required, string): How the ignition delay was
        measured, one of \yabox{d/dt max} to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox{target},
        \yabox{max} or \yabox{min} to indicate the ignition point is at the
        maximum or minimum of the \yabox{target}, or \yabox{1/2 max} to
        indicate the half-maximum point
        \item \yabox{target} (required, string): The target for the type of the
        ignition point, one of \yabox{temperature}, \yabox{pressure},
        \yabox{OH}, \yabox{OH*}, \yabox{CH}, or \yabox{CH*}
    \end{itemize}
\end{itemize}

The top-level key for the sequence is called \yabox{datapoints}. Each of the
quantities in an element must be specified with units. This should be done as a
single string associated with the first element of the sequence. The units of
the quantity are validated to ensure they have the appropriate dimensionality
for the quantity. In addition, each of the quantities in an element can
optionally be assigned an uncertainty. This uncertainty can be either absolute
or relative, and is specified as elements of the sequence of the associated key.
For example, the absolute uncertainty of the temperature and the relative
uncertainty of the ignition delay might be specified as follows:

\begin{yamlbox}
datapoints:
  - temperature:
      - 1100 kelvin
      - uncertainty-type: absolute
        uncertainty: 10 kelvin
    ignition-delay:
      - 10 us
      - uncertainty-type: relative
        uncertainty: 0.1
    ...
\end{yamlbox}

%====================================================================
\section{PyKED architecture}\label{sec:pyked-architecture}
%
\pk{} is a Python package that currently provides the reference implementation
of the interface to \ck{} files. \pk{} reads \ck{} files, validates their
structure and content, and allows the user to interact with the data contained
in the \ck{} file.

The basic user interface to the \ck{} file in \pk{} is through the
\pybox{ChemKED} class. The \pybox{ChemKED} class constructor takes the file
name of a \ck{} file or a Python dictionary containing the contents of a \ck{}
file as its argument. When the file or dictionary is loaded, it is validated to
have the correct format by the Python package Cerberus \autocite{cerberus}.

The fields of the \ck{} file are stored as instance attributes of the
\pybox{ChemKED} class. The following attributes are available:
%
\begin{itemize}
    \item \pybox{chemked_version}, \pybox{file_version}, \pybox{file_author},
    \pybox{experiment_type}: Stores the values from the \ck{} file
    \item \pybox{reference}: Instance of a \pybox{namedtuple} containing all
    the information from the literature reference associated with the data
    \item \pybox{apparatus}: Instance of a \pybox{namedtuple} containing all
    the information about the apparatus used to perform the experiment
    \item \pybox{datapoints}: A Python list of \pybox{DataPoint} instances
\end{itemize}

The \pybox{DataPoint} class stores the information associated with a single data
point in the \ck{} file (i.e., a single element of the \yabox{datapoints}
sequence). Similar to the \pybox{ChemKED} class, information in the
\pybox{DataPoint} is stored as instance attributes:
%
\begin{itemize}
    \item \pybox{equivalence_ratio}: The value of the equivalence ratio, if
    present. For informational purposes only, no validation is done of the value
    \item \pybox{ignition_delay}, \pybox{temperature}, \pybox{pressure},
    \pybox{pressure_rise}: The value of these data are stored as instances of
    the \pybox{Quantity} class from the Pint \autocite{Grecco2016} package,
    possibly with associated uncertainty
    \item \pybox{composition}: A list of dictionaries of the species and their
    respective amounts. The values are validated so that \yabox{mole-percent},
    \yabox{mole-fraction}, or \yabox{mass-fraction} cannot be mixed for a single
    data point, and so that the sum of the values is approximately 1.0, or
    100.0 for \yabox{mole-percent}.
    \item \pybox{composition_type}: A string indicating the type of composition
    information for the data point---one of \pybox{'mole-percent'},
    \pybox{'mole-fraction'}, or \pybox{'mass-fraction'}
    \item \pybox{volume_history}: If the \yabox{volume-history} of an RCM
    experiment is provided in the \ck{} file, it is stored in this attribute as
    a \pybox{namedtuple}, and the actual values are stored in NumPy arrays
    \autocite{vanderWalt:2011np}.
    \item \pybox{compression_time}: If the \ck{} file encodes an RCM experiment,
    the length of the compression stroke may be included
\end{itemize}

The \pybox{DataPoint} class has two instance methods defined, called
\pybox{get_cantera_mole_fraction()} and \pybox{get_cantera_mass_fraction()}.
With these methods, the composition of the reactant mixture can be output to a
format that can be used to set the composition of a Cantera \pybox{Solution}.
Note that the \pybox{composition} specification does not contain the molecular
weights of the components, so conversion between mole fractions and mass
fractions is not possible.

The \pybox{ChemKED} class has one instance method defined, called
\pybox{get_dataframe()}. Using this method, the user can get an instance of a
Pandas~\autocite{pandas} \pybox{DataFrame} that contains the information in the
list of \pybox{DataPoint}s. The user can specify the columns included in the
\pybox{DataFrame} by passing a list of column names to the
\pybox{output_columns} argument of the \pybox{get_dataframe()} method. The
possible columns are not case-sensitive and are:

\noindent\begin{itemize*}
    \item \pybox{'Temperature'}
    \item \pybox{'Pressure'}
    \item \pybox{'Ignition Delay'}
    \item \pybox{'Composition'}
    \item \pybox{'Equivalence Ratio'}
    \item \pybox{'Reference'}
    \item \pybox{'Apparatus'}
    \item \pybox{'Experiment Type'}
    \item \pybox{'File Author'}
    \item \pybox{'File Version'}
    \item \pybox{'ChemKED Version'}
\end{itemize*}

In addition, specific fields from the \pybox{Reference} and \pybox{Apparatus}
attributes can be included by specifying the name after a colon. These options
are:

\noindent\begin{itemize*}
    \item \pybox{'Reference:Volume'}
    \item \pybox{'Reference:Journal'}
    \item \pybox{'Reference:DOI'}
    \item \pybox{'Reference:Authors'}
    \item \pybox{'Reference:Detail'}
    \item \pybox{'Reference:Year'}
    \item \pybox{'Reference:Pages'}
    \item \pybox{'Apparatus:Kind'}
\end{itemize*}

% This weirdness is necessary to prevent the list from running over the margin
\noindent\begin{itemize*}
    \item \pybox{'Apparatus:Facility'}
    \item \pybox{'Apparatus:Institution'}
\end{itemize*}

Only the first author is printed when \pybox{Reference} or
\pybox{Reference:Authors} is selected because the whole author list may be
quite long.

@@

The \ck{} file that will be used in this example can be found in the
\pybox{tests} directory of the PyKED repository,
\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the data point is from the work of
\textcite{Mittal2006a}:
%
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, there is just a single datapoint in this file, which codes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure, compression time, ignition type, and the volume history that specifies
how the volume of the reactor varies with time, simulating the compression
stroke and post-compression processes.
%
\begin{yamlbox}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        ...
\end{yamlbox}
%
The values for the \yabox{volume-history} are truncated to save space. A user
might like to load the information from this file to perform a simulation to
calculate the ignition delay using Cantera. All of the information required to
perform this simulation is present in the \ck{} file, with the exception of a
chemical kinetic mechanism for H$_2$/CO combustion.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox{DataPoint} class. \pk{} uses Pint \pybox{Quantities} to
store values with units, while Cantera expects a floating point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_0 = dp.temperature.to('K').magnitude
P_0 = dp.pressure.to('Pa').magnitude
X_0 = dp.get_cantera_mole_fraction()

# Load the mechanism and set the initial state of the mixture
gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_0, P_0, X_0

# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox{IdealGasReactor}, a
\pybox{Wall} is installed between the reactor and the environment and assigned
a velocity. Although we do not show the details here, a reference implementation
of a class that computes a wall velocity given the volume history of the reactor
is available in CanSen~\autocite{cansen}, in the
\pybox{cansen.profile.VolumeProfile} class.
%
\begin{pythonbox}
time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

netw = ct.ReactorNet([reac])

# Integrate for 50 ms
while netw.time < 0.05:
    netw.step()
\end{pythonbox}
%
Although not shown in this example, the user would typically store information
about the \pybox{IdealGasReactor} (e.g., temperature, pressure, mass fractions)
at each time step of the integration for post-processing.

@@

The \ck{} file used in this example can be found online.\todo{add url}{} The
data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol. We have omitted the file
meta information below for space; the format is largely similar to the example in Section~\ref{sec:rcm-modeling}.
%
In this \ck{} file, multiple data points are specified with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox{common-properties} section is specified.
%
\begin{yamlbox}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
\end{yamlbox}
%
This block utilizes the ability of YAML files to define an anchor using the
\yabox{&} symbol and refer to that section later using the \yabox{*} symbol, as
can be seen in the definition of the \yabox{datapoints} section:
%
\begin{yamlbox}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{yamlbox}

In this example, we would like to run constant volume simulations at each of the
pressure and temperature conditions in the \yabox{datapoints} list. Once again,
the \ck{} file specifies all the information required for the simulations except
for the chemical kinetic mechanism.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

@@

# Since the composition is specified by the common-properties,
# just take the composition of the first DataPoint
X_0 = ck.datapoints[0].get_cantera_mole_fraction()

# Loop through each of the DataPoints
for dp in ck.datapoints:
    T_0 = dp.temperature.to('K').magnitude
    P_0 = dp.pressure.to('Pa').magnitude
    gas.TPX = T_0, P_0, X_0
    reac = ct.IdealGasReactor(gas)
    netw = ct.ReactorNet([reac])
    # Define ignition delay as T_0 + 400 K
    while reac.T < T_0 + 400:
        netw.step()

    print('The ignition delay for T_0={}, P_0={} is: {} seconds'.format(
        T_0, P_0, netw.time
    ))
\end{pythonbox}
%
Note that the ignition delay in this example is defined as equal to $T_0 +
400\,\mathrm{K}$, a simplified definition used for this example. In general, the
user could post-process the concentration information from the simulation to
determine one-half the maximum OH* concentration to match the experimental
definition of ignition delay.
%====================================================================
\section{Conclusions and future work}

Recognizing that the community has a need for an open and standardized data
serialization format, in this paper we have presented the \ck{} data format for
describing measurements taken from fundamental combustion experiments. \ck{}
files are formatted using the YAML language and are plain-text, human- and
machine-readable, and easy to construct.
